Neural Connectome of the Ctenophore Statocyst

### 2025-05-06 manuscript/Manuscript.tex
MS

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Arial}
  \setmonofont[]{DejaVu Sans Mono}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[lmargin=0.8in,rmargin=0.8in,tmargin=0.8in,bmargin=0.8in]{geometry}
\ifLuaTeX
  \usepackage{luacolor}
  \usepackage[soul]{lua-ul}
\else
  \usepackage{soul}

\fi
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

@@

\usepackage{lineno}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Cell-type diversity and connectome of the ctenophore apical organ},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Cell-type diversity and connectome of the ctenophore apical
organ}
\author{}
\date{}

@@

\setstretch{1.2}
\hfill\break

Kei Jokura\textsuperscript{1,3,*}, Sanja Jasek\textsuperscript{1,2},
Lara Kewalow\textsuperscript{2}, Pawel Burkhardt\textsuperscript{4},
Gáspár Jékely\textsuperscript{1,2,*}\\

\textsuperscript{1}Living Systems Institute, University of Exeter,
Exeter, EX4 4QD, United Kingdom\\
\textsuperscript{2}Heidelberg University, Centre for Organismal Studies
(COS), 69120 Heidelberg, Germany\\
\textsuperscript{3}Exploratory Research Center on Life and Living
Systems (ExCELLS), National Institutes of Natural Sciences, Okazaki,
Japan\\
\textsuperscript{4}Michael Sars Centre, University of Bergen, Norway\\
\textsuperscript{*}Correspondence: jokura@nibb.ac.jp,
gaspar.jekely@cos.uni-heidelberg.de

\subsection{Abstract}\label{abstract}

\textbf{text in bold} \emph{italic} \ul{underline}

@@

First descriptions of AO: (Aronova, 1974)

The lobate ctenophore \emph{Mnemiopsis}

The lithocytes have a large calcium carbonate concretion and\ldots{}
(Tamm, 2014a) Living lithocytes are transported along the balancer cilia
to the tip of the statocyst (Noda and Tamm, 2014) The projections of a
group of bridge cells located on two sides of the apical organ forms an
arch and connects the balancers along the tentacular plane (Tamm and
Tamm, 2002).

The easiest way is to use the command line to add new refs to the .bib
file:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#curl {-}LH "Accept: application/x{-}bibtex" https://doi.org/10.7554/eLife.91258.1 \textgreater{}\textgreater{} references.bib}
\end{Highlighting}
\end{Shaded}

@@

\subsubsection{\texorpdfstring{Volume EM reconstruction of the
\emph{Mnemiopsis} apical
organ}{Volume EM reconstruction of the Mnemiopsis apical organ}}\label{volume-em-reconstruction-of-the-mnemiopsis-apical-organ}

\begin{itemize}
\tightlist
\item
  specimen
\item
  volume, resolution, fixation, quality
\item
  number of sections
\item
  skeletonisation in Catmaid
\item
  number of cells
\item
  general anatomy, regions, quadrants
\end{itemize}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure1.png}

@@

\caption{\textbf{Figure 1. Title fig 1} pic of a larva, boxed region of
AO, schematic, catmaid pic with slice, dimensions, quadrants, regions,
(A) legend (B) legend.}

@@

\subsubsection{Classification of cell
types}\label{classification-of-cell-types}

\begin{itemize}
\tightlist
\item
  ultrastructural criteria (cilia, vesicles etc)
\item
  number of types, cells per type - table
\end{itemize}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure_network.png}

@@

\caption{\textbf{Figure 2. Title fig 2} Balancer complex: balancer,
lithocyte, groove 2 types, bridge - EM pics, (A) legend (B) legend.}

@@

``the larval bridge has not yet developed forks at its ends, as seen in
adult Mnemiopsis bridges''(Tamm and Tamm, 2002)

Figure3 - dome, bicil, monocil, no-cil, plumose, dense vesicle cells
(rename), gland? cells, oligo/multiciliated cells

Plumose cells (Hernandez-Nicaise, 1984) may be pressure sensors (Tamm,
2014b)

Figure4 - lamellate body complex, LB, intra-multi-ciliated, relationship
to balancer

Figure5 - nerve net, different fragments/cells/syntytia, EM, relation to
entire organ

Figure 6 - cilia? renderings of basal bodies, centrioles, length,
axoneme structure (table), table about cell types, number of cilia etc.

\subsection{Synapses}\label{synapses}

``synaptic regions with the characteristic presynaptic triad mor-
phology of ctenophore nerves (i.e., a single layer of vesicles, smooth
ER sac, and closely apposed mito- chondria (Hernandez-Nicaise, 1973;
Horridge and Mackay, 1964) Hernandez- Nicaise, 1973)''

Figure 7 - synapses, reconstruction, annotation, numbers, distribution
of partners, example EM (figure supplement - lots of EM), mitochondria
(Fig suppl - mitochondrion stats - Sanja)

Gap juntions? Innexins are expressed in the apical organ in specific
patches of cells in each quadrant (Ortiz et al., 2023)

Ctenophore synapses (Hernandez-Nicaise, 1973)

We found no synapses between bridge and balancer, in agreement with Tamm
(Tamm and Tamm, 2002).

Figure 8 - connectome cell-type based, quadrant based, nerve net
syntycia, bridge, quadrants (Fig suppl - cell-based network)

Tamm\&Tamm observed synapses from neurites to bridge cells (2002). and
in Mnemiopsis larvae observed synapses from bridge cells onto other
cells

Inserting Figures

>>>>

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{Arial}
    \setmonofont[]{DejaVu Sans Mono}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[lmargin=0.8in,rmargin=0.8in,tmargin=0.8in,bmargin=0.8in]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\usepackage{lineno}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Neural Connectome of the Ctenophore Statocyst},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Neural Connectome of the Ctenophore Statocyst}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\textbf{Kei Jokura}\(^{1,2,3,4,5*}\), \textbf{Sanja
Jasek}\(^{1,2,6}\), \textbf{Lara Kewalow}\(^{6}\), \textbf{Pawel
Burkhardt}\(^{7}\), \textbf{Gáspár Jékely}\(^{1,2,6,*}\) \(^{1}\) Living
Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom
\(^{2}\) Biosciences, Faculty of Health and Life Sciences, University of
Exeter, Exeter EX4 4QD, UK \(^{3}\) Grass Laboratory, Marine Biological
Laboratory, Woods Hole, MA 02543, USA \(^{4}\) Exploratory Research
Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan
\(^{5}\) National Institute for Basic Biology (NIBB), Okazaki, 444-8585,
Japan \(^{6}\) Heidelberg University, Centre for Organismal Studies
(COS), 69120 Heidelberg, Germany \(^{7}\) Michael Sars Centre,
University of Bergen, Norway \textbf{\(^{*}\) Correspondence:}
\href{mailto:jokura@nibb.ac.jp}{\nolinkurl{jokura@nibb.ac.jp}},
\href{mailto:gaspar.jekely@cos.uni-heidelberg.de}{\nolinkurl{gaspar.jekely@cos.uni-heidelberg.de}}}
\author{}
\date{}

@@

\setstretch{1.2}
\subsection{Abstract}\label{abstract}

Ctenophores possess a unique gravity receptor (statocyst) in their
aboral organ formed by four clusters of ciliated balancer cells that
collectively support a statolyth. During reorientation, differential
load on the balancer cilia leads to altered beating of the ciliated comb
rows to elicit turns. To study the neural bases of gravity sensing, we
imaged by volume electron microscopy (vEM) the aboral organ of the
ctenophore \emph{Mnemiopsis leidyi}. We reconstructed 909 cells,
including syncytial neurons that form a nerve net. The syncytial neurons
synapse on the balancer cells and also form reciprocal connection with
the bridge cells that span the statocyst. High-speed imaging revealed
that balancer cilia beat and arrest in a coordinated manner but with
differences between the sagittal and tentacular planes of the animal,
reflecting nerve-net organisation. Our results suggest a coordinating
rather than effector function for the nerve net and inform our
understanding of the diversity of nervous-systems organisation across
animals.

@@

Ctenophores, or comb jellies, are gelatinous marine animals that
actively swim by beating rows of fused cilia known as comb plates. These
coordinated ciliary movements not only generate propulsion but also
allow precise control of body orientation and directional changes within
the water column. Despite lacking a centralized nervous system,
ctenophores exhibit sophisticated and dynamic behavioral patterns and
respond to a broad range of external stimuli, suggesting the presence of
an evolutionarily unique and functionally integrated sensorimotor
system. Several components of the ctenophore nervous system have been
described, including a subepithelial nerve net (SNN), mesogleal neurons,
tentacular nerves, and elements of nervous system in the aboral organ
(Hernandez-Nicaise, 1984; Hertwig, 1880; Jager et al., 2010). These
neural structures are characterized by unique features such as a
syncytial architecture (Burkhardt et al., 2023) and an extensive
repertoire of lineage-specific neuropeptides not found in other
metazoans (Hayakawa et al., 2022; Sachkova et al., 2021). Notably, early
single-cell transcriptomic studies failed to identify canonical neuronal
clusters using conventional markers (Sebé-Pedrós et al., 2018). It was
only through the use of pro-neuropeptide-based markers that distinct
neural cell types were subsequently identified (Hayakawa et al., 2022;
Sachkova et al., 2021). In parallel, phylogenomic and chromosomal
synteny analyses have suggested that ctenophores may represent the
sister group to all other animals (Li et al., 2021; Ryan et al., 2013;
Schultz et al., 2023; Whelan et al., 2015), raising the possibility that
their nervous system evolved independently from those of other metazoans
(Moroz, 2015). This hypothesis positions ctenophores as a key lineage
for re-evaluating the origins and diversity of neural systems. However,
without a detailed understanding of how ctenophore neurons are spatially
organized and interconnected, it remains unclear how their nervous
system gives rise to behavior. Clarifying this relationship between
neural architecture and behavioral control is essential to uncovering
the logic of ctenophore neural function.

To address the lack of structural understanding of how ctenophores
control behavior, we focused on a unique sensory structure---the aboral
organ. This gravity-sensing organ contains a statocyst supported by four
mechanosensory balancer cells, which act as pacemakers for comb plate
beating and posture control (HORRIDGE, 1965; Tamm, 1980).
Ultrastructural and immunohistochemical studies have revealed that the
aboral organ contains diverse sensory and neural components, including a
distinct neural structure termed the deep nerve net (Aronova, 1974;
HERNANDEZ-NICAISE, 1968; Jager et al., 2010). A key feature of balancer
cilia is their bidirectional, mechanoresponsive control of beat
frequency. Depending on the direction of deflection and the animal's
geotactic state, balancer cilia may either increase or decrease their
activity, indicating a dynamic response system not fixed like vertebrate
hair cells (HORRIDGE, 1965; Tamm, 1982; Tamm, 2014a). Previous studies
have implicated membrane potential, intracellular calcium concentration,
and synaptic input in modulating this response (Lowe, 1997). In
particular, deflection-induced excitation can be abolished by calcium
channel inhibitors or calcium-free conditions (Lowe, 1997), while high
extracellular K⁺ depolarizes and activates cilia in the absence of
mechanical stimuli (Lowe, 1997). Local calcium application further
suggests that Ca²⁺ influx at the ciliary base may be required for the
excitatory response (Lowe, 1997). These findings support the hypothesis
that neural modulation plays a critical role in shaping balancer cell
output (Lowe, 1997).

To investigate how balancer cilia are modulated within the aboral organ,
we combined volume electron microscopy and high-speed imaging in the
ctenophore \emph{Mnemiopsis leidyi}. Through connectomic reconstruction,
we characterized the synaptic architecture of the aboral nerve net, and
functional imaging revealed correlated activity patterns in balancer
cilia. These results suggest that the aboral nerve net plays a central
role in integrating sensory input and coordinating motor output.

@@

\subsubsection{\texorpdfstring{Volume EM reconstruction of the
\emph{Mnemiopsis} aboral
organ}{Volume EM reconstruction of the Mnemiopsis aboral organ}}\label{volume-em-reconstruction-of-the-mnemiopsis-aboral-organ}

For vEM analysis, we used a five-day-old cydippid larva of
\emph{Mnemiopsis leidyi}. To obtain high-quality ultrastructural
preservation, we used high pressure freezing followed by freeze
substitution and resin embedding (Epon). The specimen was sectioned from
the aboral tip and we collected approximately 1,000 ultra-thin (50 nm)
serial sections as ribbons. The sections were imaged in a Zeiss Gemini
500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only
the region containing the aboral organ. Subsequently, 620 of these
images were stitched and aligned in TrakEM2 (Cardona et al., 2012). The
final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We
skeletonised and annotated all cells in this dataset in CATMAID
(Saalfeld et al., 2009), resulting in a reconstruction of 909 cells with
nuclei and cell bodies intact. Most of the cells, including the
balancers and the aboral nerve net neurons are fully contained within
the volume, due to the compact and mostly self-contained organisation of
the aboral organ.

In cells containing cilia, we also traced cilia along their length and
annotated all basal bodies. For neuronal skeletonization, nodes were
placed to interconnect the profiles of the same neuron's processes
across layers, extending the skeleton until all branches were fully
traced. All nodes relevant to synaptic sites were tagged, and skeletons
were named and assigned multi-level annotations. As described later,
some neurons formed loop-like structures (anastomosed neurons), wherein
separated branches often rejoined either the main branch or smaller
branches (Burkhardt et al., 2023). In such cases, branch nodes were
placed near the closest existing node and annotated accordingly (CATMAID
only supports skeleton trees). The entire skeletonized volume was
composed of 134,591 nodes. 88 fragments could not be attached to
somata-associated skeletons. Most of these fragments represent short
skeletal branches that could not be traced beyond gaps or low-quality
layers.

Next, we divided the entire aboral organ into four broad quadrants
(Q1-Q4) to facilitate grouping the identified cells (Figure 1G)
(Martindale and Henry, 1999). The general body plan of ctenophores, when
viewed from the aboral side, exhibits biradial symmetry around the anal
pores (Martindale and Henry, 1999). This symmetry corresponds to the
four blastomeres present at the four-cell stage during early embryonic
development (Martindale and Henry, 1999). We assigned each of the 909
skeletonised cells to one of the four quadrants.

@@

{\centering \includegraphics{figures/Figure1.png}

@@

\caption{\textbf{Figure 1. Morphology of the aboral organ in
\emph{Mnemiopsis leidyi}}\\
(A) Whole-body image of a 5-day-old \emph{M. leidyi} cydippid larva in
lateral view. The boxed region indicates the aboral organ (ao). Scale
bar: 100 µm.\\
(B) Schematic diagram of the aboral view of a cydippid larva. The
two-tone coloration represents the biradial symmetry of the body. For
convenience, the body is divided into four quadrants, designated as the
first (Q1) through fourth (Q4) quadrants. The two primary viewing angles
are referred to as the sagittal plane (S) and the tentacular plane (T).
Abbreviations: ao, aboral organ; cg, ciliated groove; cr, comb plates;
tb, tentacle bulbs.\\
(C) Aboral views of the aboral organ highlighting its spatial
organization. The left panel presents a schematic representation of the
aboral organ, illustrating the four body quadrants (Q1--Q4), color-coded
consistently with panel (B). The right panel shows a corresponding
differential interference contrast (DIC) image from the same
perspective. The region enclosed by the dotted line delineates the
boundary of the aboral organ. The sagittal (S) and tentacular (T) arrows
indicate the viewing directions corresponding to those planes. Scale
bar: 50 µm. bal, balancer; cg, ciliated grooves; li, lithocyte.\\
(D) Lateral views of the aboral organ in two orthogonal planes captured
using DIC microscopy. The left panel shows a lateral view in the
sagittal (S) plane, while the right panel displays a view in the
tentacular (T) plane. Balancer cilia (bal) and lithocytes (li) are
enclosed within the dome (do). Scale bar: 50 µm. A--O indicates the
aboral--oral body axis.\\
(E) Schematic representations of the aboral organ in sagittal and
tentacular planes. The left panel shows a lateral schematic in the
sagittal plane (S), and the right panel shows the same in the tentacular
plane (T), corresponding to the views in panel (D). Colors match the
quadrant scheme shown in panels (B) and (C).\\
(F) Example of cell tracing using the collaborative annotation toolkit
CATMAID for large-scale electron microscopy image datasets. The
spherical objects indicate nuclear positions, while the lines represent
the traced cell centers. Scale bar: 10 µm.\\
(G) Three-dimensional reconstruction of cells composing the aboral
organ, displayed from different perspectives. The left panel presents an
aboral view of the aboral organ. The middle panel shows a sagittal plane
view, while the right panel provides a tentacular plane view. Cells are
color-coded according to their respective quadrants. Lithocytes (li) are
represented as three gray spheres, while balancers (bal) are depicted as
gray lines. Scale bar: 25 µm.}

@@

\subsubsection{An aboral synaptic nerve net of syncytial
neurons}\label{an-aboral-synaptic-nerve-net-of-syncytial-neurons}

Classical neural staining techniques did not provide clear images of
neurons at the aboral organ. However, ultrastructural studies revealed
morphological evidence for neural elements, including synapses that are
located on the epithelial floor of the aboral organ---a region where
epithelial cells are densely packed and form a floor-like structure
(Hernandez-Nicaise, 1973; Horridge and Mackay, 1964). In our dataset, we
identified synapses based on the previously described pre-synaptic triad
morphology consisting of a single layer of vesicles, a cisterna of
smooth endoplasmic reticulum and an associated mitochondrion. At
synaptic sites, we marked mitochondria as synaptic nodes (orange in
CATMAID) and connected this node to the nearest node in postsynaptic
cells across the regions where synaptic vesicles aligned with the
presynaptic cell's membrane (light blue arrows in CATMAID)
(Supplementary Figure S2A). We could not detect specialized
post-synaptic structures, in agreement with previous studies
(Hernandez-Nicaise, 1973). Synapses were identified as either monoadic
or polyadic, with one pre-synaptic neuron forming connections with one
or multiple post-synaptic cells (Supplementary Figure S2B).

We reconstructed three aboral nerve-net (ANN) neurons, each with a
syncytial morphology (Figure 2A and B). Each ANN had multiple nuclei
(ranging from two to five per neuron), with anastomosing membranes
(Supplementary Figure S2C). These neurons are distinct from the
syncytial subepithelial nerve net (SNN) neurons with blebbed morphology
previously identified by vEM in the body wall of \emph{Mnemiopsis}
(Burkhardt et al., 2023). Our results reveal a second type of syncytial
neurons in ctenophores (Supplementary Figure S2D).

Furthermore, ANNs are also different from other neurons previously
reported by EM, including mesogleal neurons and ciliated sensory cells
that synapse on the SNN (types 1-4)(Burkhardt et al., 2023). Based on
their distinct size and position, we classified the ANNs into two types.
The first type is a single large neuron (ANN\_Q1-4) spans all four
quadrants (Q1-Q4) and has four (possibly five) nuclei. This neuron has
150 presynaptic sites. The second type consists of two neurons
(ANN\_Q1Q2 and ANN\_Q3Q4), each spanning two quadrants and containing
two nuclei. ANN\_Q1Q2 had 119 presynaptic structures, whereas ANN\_Q3Q4
had 68 presynaptic structures.

@@

{\centering \includegraphics{figures/Figure2.png}

@@

\caption{\textbf{Figure 2. Organisation of the aboral synaptic nerve
net}\\
(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4),
which spans all four quadrants and contains multiple nuclei (blue). The
spheres represent the positions of individual nuclei. The left panel
shows a aboral view of the aboral organ, the middle panel presents a
sagittal plane view, and the right panel provides a tentacular plane
view.\\
(B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each
spanning two quadrants and containing multiple nuclei. The spheres
indicate the positions of the nuclei. The left panel shows a aboral view
of the aboral organ, the middle panel presents a sagittal plane view,
and the right panel provides a tentacular plane view.\\
(C) Graph showing the number of mitochondria per cell and the proportion
of mitochondria involved in synapse formation for each cell type. Green
represents mitochondria associated with synapses, while gray represents
mitochondria not involved in synapse formation. ANN has a significantly
higher number of mitochondria compared to other cell types, with most of
them forming synapses. Except for lithocytes, each cell type has a
sample size of at least three cells. The abbreviations are as follows:
bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv
(dense vesicle cells), imc (intra-multiciliated cells), la (lamellate
bodies), li (lithocytes), pl (plumose), and ef (epithelial floor
cells).\\
(D) Localization of mitochondria within ANN Q1-4. Red indicates
mitochondria associated with the presynaptic triad structures, yellow
marks mitochondria containing synaptic vesicles but lacking a clearly
defined presynaptic triad, blue represents mitochondria with unclear
synaptic vesicles, and black denotes mitochondria where no synaptic
vesicles were identified.\\
(E) Representative electron micrographs of presynaptic triad structures
observed in the dataset and their schematic diagrams. The left diagram
illustrates synaptic projections from the center ANN (ANN Q1-4) to the
lateral ANN (ANN Q3Q4), while the right diagram shows synaptic
projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN
Q1-4). The abbreviations are as follows: mi (mitochondrion), er
(endoplasmic reticulum), and sv (synaptic vesicles).\\
(F) Plot showing the three-dimensional positions of synapses. Synapses
from the center ANN to the lateral ANN are shown in magenta, while
synapses from the lateral ANN to the center ANN are shown in cyan. Blue
dots indicate the locations of autapses within ANN Q1-4.}

@@

\subsubsection{Synaptic connectome of the gravisensory
organ}\label{synaptic-connectome-of-the-gravisensory-organ}

Our synapse annotation revealed synapses between the ANN neurons and
synapses that ANN\_Q1-4 forms on itself (autapses). The ANN neurons also
form synapses on the gravity-sensing balancer cells.

Balancer cells are monociliated cells with long motile cilia that form
four bundles of compound cilia, one in each quadrant. These bundles come
together at the centre of the aboral organ to support the cellular mass
of the statolyth. The bending of the cilia and the position of somata
differed clearly when viewed laterally from the sagittal or the
tentacular plane (Figure 3B). In each quadrant, there were between 28-37
balancer cells (Q1: 37; Q2: 32; Q3: 32; Q4: 28). Each cell contained 3
to 10 mitochondria.

ANN\_Q1-4 formed synapses on balancer cells in all four quadrants (on 7
cells in Q1, 11 cells in Q2, 6 cells in Q3 and 10 cells in Q4) while
ANN\_Q1Q2 and ANN\_Q3Q4 synapsed to balancer cells in their respective
quadrants (ANN\_Q1Q2 on 6/37 cells in Q1 and 8/32 cells in Q2; ANN\_Q3Q4
on 1/32 cells in Q3 and 5/28 cells in Q4). Some balancer cells received
inputs from both ANN\_Q1-4 and either of ANN\_Q1Q2 or ANN\_Q3Q4. While
previous studies suggested the presence of afferent synapses from
balancer cells to neurons (Hernandez-Nicaise, 1974), our data did not
reveal any presynaptic sites in balancer cells.
